---
title: "LUAD"
author: "Cihan"
date: "27 12 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DATA

### Load the Data
```{r}
missing.values <- c("N/A","na","?","", "NaN", "not reported")
data <- read.csv("C:\\Users\\sakma\\Desktop\\Ders Kayıtları\\Data Mining\\Final Project\\TCGA-LUAD_clinical.csv", na=missing.values)
```

### **We should give some information about data....**
```{r cars}
summary(data)
```

# PREPROCESSING

## First we're going to drop the fully NaN columns. Because these columns are not useful for us.
Maybe we can show these fully nan columns in a fashionable way.

**we are going to get a rough glimpse on the missingness situation with the pretty neat naniar package**

**Note:** we can go through https://towardsdatascience.com/smart-handling-of-missing-data-in-r-6425f8a559f2 and leave some comments for our dataset.
```{r}
#install.packages("naniar")
library(naniar)

# Are there missing values in the dataset?
any_na(data)
# How many?
n_miss(data)
prop_miss(data)
# Which variables are affected?
data %>% is.na() %>% colSums()


#data_new <- data[ , colSums(is.na(data)) < nrow(data)]
#summary(data_new)

```





























