---
title: "LUAD"
author: "Cihan"
date: "27 12 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DATA

### Load the Data
```{r}
missing.values <- c("N/A","na","?","", "NaN", "not reported", "NA")
data <- read.csv("C:\\Users\\sakma\\Desktop\\Ders Kayıtları\\Data Mining\\Final Project\\TCGA-LUAD_clinical.csv", na.strings=missing.values)
```

### **We should give some information about data....**
```{r cars}
summary(data)
```

# PREPROCESSING

## First we're going to drop the fully NaN columns. Because these columns are not useful for us.
Maybe we can show these fully nan columns in a fashionable way.

**we are going to get a rough glimpse on the missingness situation with the pretty neat naniar package**

**Note:** we can go through https://towardsdatascience.com/smart-handling-of-missing-data-in-r-6425f8a559f2 and leave some comments for our dataset.
```{r}
#install.packages("naniar")
library(naniar)
library(ggplot2)

# Are there missing values in the dataset?
any_na(data)
# How many?
n_miss(data)
prop_miss(data)
# Which variables are affected?
data %>% is.na() %>% colSums()


data_new <- data[ , colSums(is.na(data)) < nrow(data)]
summary(data_new)

```

Now let's take a closer look at to the missing values with tables.
```{r}
# Get number of missings per variable (n and %)
miss_var_summary(data_new)
#miss_var_table(data)

```

We can see the missing values with descending order in a more visualize version.
```{r}
# Which variables contain the most missing variables?
gg_miss_var(data_new)
```

To get a better understanding whether or not the data are missing at random, we are going to visualize the locations of missing values across all variables.
```{r}
# Where are missings located?
vis_miss(data_new) + theme(axis.text.x = element_text(angle=80))
```

**There is a direct realtionship between *days_to_last_follow_up* and *days_to_death* columns**  
- In *days_to_death* column values are NaN for **alive** patients and there is just information about **dead** patients how long they lived.

























































